### 2022-09-01 main/chapter_introduction.tex
add least squares note; add labels note; fix pa pb independent

The `which is real?' method is general, and can test not only parity, but
symmetry under any transformation of the data, irrespective of whether it is
discrete or continuous.
In the context of modern machine learning, it is an example of
`self-supervised' learning, in which the training objective derives from the
data and not from additional labels~\cite{
multitaskself2017,
devlin2019bert
}, and quite similar to the `CLIP' objective, which also attempts to identify
a real examples from collections of alternatives~\cite{pmlr-v139-radford21a}.
What is special in our work is our application for scientific self-supervised
testing.

>>>>

The `which is real?' method is general, and can test not only parity, but
symmetry under any transformation of the data, irrespective of whether it is
discrete or continuous.
In the context of modern machine learning, it is an example of
`self-supervised' learning, in which the training objective derives from the
data and not from additional labels~\cite{
Noroozi2016jigsaw,
multitaskself2017,
devlin2019bert
}, and quite similar to the `CLIP' objective, which also attempts to identify
a real examples from collections of alternatives~\cite{pmlr-v139-radford21a}.
What is special in our work is our application for scientific self-supervised
testing.

### 2022-09-01 main/chapter_searches.tex
add least squares note; add labels note; fix pa pb independent

Statistical data analysis suffers from dogma --- prescriptions and recipes that
students are taught to memorize and follow without knowing their reasons or
limitations.
From a standard undergraduate physics education, my experience has been
error propagation formulae,
least-squares fitting routines,
chi-squared per degree of freedom,
and myriad statistical tests and confidence intervals.
Many of these are effective approximations, but all also fail if misused
outside their domain.%
\footnote{%
Multiplicative error propagation is an example.
For $a = \mu_a \pm \sigma_a$ and $b = \mu_b \pm \sigma_b$,
the standard perturbative approximation is
\(
\widetilde{\sigma}_{ab}
= \mu_a \mu_b \sqrt{\sigma_a^2/\mu_a^2 + \sigma_b^2/\mu_b^2}
% = \sigma_a^2 \mu_b^2 + \sigma_b^2 \mu_a^2
\),
but a short calculation with their moments derives the exact result
\(
\sigma_{ab}
= \sqrt{(\mu_a^2 + \sigma_a^2)(\mu_b^2 + \sigma_b^2) - \mu_a^2\mu_b^2}
% = \sigma_a^2 \mu_b^2 + \sigma_b^2 \mu_a^2 + \sigma_a^2 \sigma_b^2
\)
if $\probc{a}$ and $\probc{b}$ are independent.
Since $\sigma_{ab}^2 = \widetilde{\sigma}_{ab}^2 + \sigma_a^2 \sigma_b^2$, the
propagation formula works only for small relative errors.
}
All effective methods of data analysis can be derived upwards from the roots of
probability and decision theory, and as elsewhere in physics, by knowing the
derivations we can also know when each approximation should work and when it
needs replacing.

@@

The Gaussian distribution has many justifications:
as the unique distribution that is spherically symmetric and factors linearly
into independent dimensions~\cite{
jaynes2003probability,
herschel1850normal,
maxwell1860normal,
muller1959note,
marsaglia1972normal
},
as a practical approximation that permits analysis in linear algebra,
and as the maximum entropy~\cite{PhysRev.106.620} distribution for a
known mean and variance.
This maximum entropy statement means that for a distribution with known mean
and variance in given coordinates, its best Gaussian approximation matches
those moments.\footnote{%
By `best', this statement means a minimal Kullback-Leibler divergence
\(
H(p\leftarrow q) = \sum_i p_i \log(p_i/q_i)
\)
from the Gaussian approximation $q$ to the target $p$ with known moments.
Such an approximation may not be \emph{good}, but it is best in class.%
}
As a limiting approximation, the Gaussian form arises from products of many
functions, such as combinations of many independent measurements that multiply
their likelihood functions.
At a maximum $\check{x}$, the first derivative of a function vanishes and the
second derivative is negative.
Taylor expanding about this maximum to the leading order, the product looks like
\begin{equation}
f(x)^N = \displaystyle \lim_{N \to \infty}
\left(
1 - \kappa (x - \check{x})^2
\right)^N,
\end{equation}
which, again by Equation~\ref{eqn:searches_exponential}, goes into the Gaussian
form
\begin{equation}
\label{eqn:searches_gaussian}
\mathrm{Gaussian}(x\mid \mu, \sigma)
\,\mathrm{d}x =
\frac{1}{\sqrt{2\pi \sigma^2}}
\exp\!\left(-\frac{1}{2}\frac{(x - \mu)^2}{\sigma^2}\right)
\mathrm{d}x
\end{equation}
where the prefactor is for normalization to a probability distribution over
$x$, and we include the $\mathrm{d}x$ terms as reminders that it is a density
function in that coordinate.

@@

In the current year, it would be remiss to not mention machine learning
algorithms.
Among various other applications, learning algorithms demonstrate practical
successes in probability assignments, such as assigning
probabilities to labels for classification,
or to assign probabilities to words for text
generation, or any of myriad other
predictions~\cite{MurphyKevinP.2012Mlap, radford2019language}.
Machine learning tunes flexible functions by consuming data and testing their
predictions in the manner we have described; their training and testing works
roughly as follows:
\begin{enumerate}
\item Predict new data with $\prob{d_i}{f(d_{i-1}, \ldots), h}$.
Keep this evidence as a testing score for comparison against other models.
(Practically, keep its equivalent $\log$ mean over independent samples.)
\item Use those data to update your state
$f(d_{i-1}, \ldots) \rightarrow f(d_i, d_{i-1}, \ldots)$ to improve your
predictions.
Unlike ideal posterior updating, this encodes only a function of the data, not
its whole --- one usually cannot afford to consume all aspects of data, so
although theoretically suboptimal this approximation is vastly superior in
practice. Learning only part of the data is not wrong, just less precise
than might be possible with a divinely perfect learning algorithm.
\item Go to 1.
\end{enumerate}
These sequential data $d_i$ might be labelled the `training', `validation',
and `testing' sets, but other splittings are possible.
This perspective highlights the importance of independent training and testing
sets --- since $\prob{d}{d, \ldots} = 1$, if any memory of $d_i$ is kept in
the $f(\ldots, d_i, \ldots)$ state, then it is a memory and not a prediction.
Independent testing data permit the evidence interpretation of predictive
probabilities.

>>>>

Statistical data analysis suffers from dogma --- prescriptions and recipes that
students are taught to memorize and follow without knowing their reasons or
limitations.
From a standard undergraduate physics education, my experience has been
error propagation formulae,
least-squares fitting routines,
chi-squared per degree of freedom,
and myriad statistical tests and confidence intervals.
Many of these are effective approximations, but all also fail if misused
outside their domain.%
\footnote{%
Multiplicative error propagation is an example.
For $a = \mu_a \pm \sigma_a$ and $b = \mu_b \pm \sigma_b$,
the standard perturbative approximation is
\(
\widetilde{\sigma}_{ab}
= \mu_a \mu_b \sqrt{\sigma_a^2/\mu_a^2 + \sigma_b^2/\mu_b^2}
% = \sigma_a^2 \mu_b^2 + \sigma_b^2 \mu_a^2
\),
but a short calculation with their moments derives the exact result
\(
\sigma_{ab}
= \sqrt{(\mu_a^2 + \sigma_a^2)(\mu_b^2 + \sigma_b^2) - \mu_a^2\mu_b^2}
% = \sigma_a^2 \mu_b^2 + \sigma_b^2 \mu_a^2 + \sigma_a^2 \sigma_b^2
\)
if $a$ and $b$ are independent.
Since $\sigma_{ab}^2 = \widetilde{\sigma}_{ab}^2 + \sigma_a^2 \sigma_b^2$, this
propagation formula works only for small relative errors.
}
All effective methods of data analysis can be derived upwards from the roots of
probability and decision theory, and as elsewhere in physics, by knowing the
derivations we can also know when each approximation should work and when it
needs replacing.

@@

The Gaussian distribution has many justifications:
as the unique distribution that is spherically symmetric and factors linearly
into independent dimensions~\cite{
jaynes2003probability,
herschel1850normal,
maxwell1860normal,
muller1959note,
marsaglia1972normal
},
as a practical approximation that permits analysis in linear algebra,
and as the maximum entropy~\cite{PhysRev.106.620} distribution for a
known mean and variance.
This maximum entropy statement means that for a distribution with known mean
and variance in given coordinates, its best Gaussian approximation matches
those moments.\footnote{%
By `best', this statement means a minimal Kullback-Leibler divergence
\(
H(p\leftarrow q) = \sum_i p_i \log(p_i/q_i)
\)
from the Gaussian approximation $q$ to the target $p$ with known moments.
Such an approximation may not be \emph{good}, but it is best in class.%
}
As a limiting approximation, the Gaussian form arises from products of many
functions, such as combinations of many independent measurements that multiply
their likelihood functions.
At a maximum $\check{x}$, the first derivative of a function vanishes and the
second derivative is negative.
Taylor expanding about this maximum to the leading order, the product looks like
\begin{equation}
f(x)^N = \displaystyle \lim_{N \to \infty}
\left(
1 - \kappa (x - \check{x})^2
\right)^N,
\end{equation}
which, again by Equation~\ref{eqn:searches_exponential}, goes into the Gaussian
form
\begin{equation}
\label{eqn:searches_gaussian}
\mathrm{Gaussian}(x\mid \mu, \sigma)
\,\mathrm{d}x =
\frac{1}{\sqrt{2\pi \sigma^2}}
\exp\!\left(-\frac{1}{2}\frac{(x - \mu)^2}{\sigma^2}\right)
\mathrm{d}x
\end{equation}
where the prefactor is for normalization to a probability distribution over
$x$, and we include the $\mathrm{d}x$ terms as reminders that it is a density
function in $x$, not a probability exactly.
On a $\log$ scale, the Gaussian form is simply quadratic in $\mu$ (or $x$);
that quadratic form continues in its multivariate generalization, and
justifies least-squares methods.

@@

In the current year, it would be remiss to not mention machine learning
algorithms.
Among various other applications, learning algorithms demonstrate practical
successes in probability assignments, such as assigning
probabilities to labels for classification,
or to assign probabilities to words for text
generation, or any of myriad other
predictions~\cite{MurphyKevinP.2012Mlap, radford2019language}.
Machine learning tunes flexible functions by consuming data and testing their
predictions in the manner we have described; their training and testing works
roughly as follows:
\begin{enumerate}
\item Predict new data with $\prob{d_i}{f(d_{i-1}, \ldots), h}$.
Keep this evidence as a testing score for comparison against other models.
(Practically, keep its equivalent $\log$ mean over independent samples.)
These $d_i$ may be only reductions of the data, such as class labels
or index permutations~\cite{Noroozi2016jigsaw, multitaskself2017}.
Other inputs that are not predicted, such as associated images that the machine
uses to predict the labels, are included in the context $h$.
\item Use those data to update your state
$f(d_{i-1}, \ldots) \rightarrow f(d_i, d_{i-1}, \ldots)$.
Unlike ideal posterior updating, this encodes only a function of the data, not
its whole --- one usually cannot afford to consume all aspects of data, so
although theoretically suboptimal this approximation is vastly superior in
practice.
Learning only part of the data is not wrong, it is just less precise than a
divinely perfect learning algorithm, which we do not currently possess.
\item Go to 1.
\end{enumerate}
These sequential data $d_i$ might be labelled the `training', `validation',
and `testing' sets, but other splittings are possible.
This perspective highlights the importance of independent training and testing
sets --- since $\prob{d}{d, \ldots} = 1$, if any memory of $d_i$ is kept in
the $f(\ldots, d_i, \ldots)$ state, then it is a memory and not a prediction.
Independent testing data permit the evidence interpretation of predictive
probabilities.
